Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-05-27 docs/submission/template.tex

\subsubsection{Surface physiographic fields}

@@

To expand V15 and V20 lake description (to V15X and V20X respectively, see Table \ref{tab:datasources}) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) would really benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant changes in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover GLDBv3 salt lake list was used. First, in order to identify separate lakes on ~1km resolution lake cover (here max lake distribution based on monthly varying lake cover is used to ease further application), small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data.

@@

 \noindent VESPER was trained on selected atmospheric and surface model fields from ERA5 (see Table \ref{table:definitions}) and Aqua-MODIS LST for 2016 and certain static version of the surface physiographic fields (see Table \ref{tab:datasources}), once VESPER was fully trained it was used to predict LST over the whole globe for 2019. Finally VESPER’s predictions can be compared to initial ERA5 skin temperatures and actual Aqua-MODIS LST for 2019 (see Figure \ref{fig:example_model}; VESPER trained using V15 surface physiographic fields is called VESPERV15, for full explanation see Table \ref{tab:vesper_table}). Figure \ref{fig:example_model} shows mean absolute errors (MAE) of differences between VESPERV15 LST predictions and Aqua-MODIS LST (see Figure \ref{fig:wasserstein_temperature}), and between ERA5 skin temperature and Aqua-MODIS LST (see Figure \ref{fig:wasserstein_precipitation}); based on that a conclusion can be drawn that VESPERV15 was able to learn corrections to ERA5, especially in the Himalayas and sub-Saharan Africa as well as Australia and the Amazon basin, leading to the globally averaged MAE reduction for predicted LST (MAEERA5=3.9K comparing to MAEVESPERV15=3.0K). All VESPER versions are trained with ERA5 fields for 2016 and with main surface physiographic fields V15. Then depending on the version some or all additional surface physiographic fields (see Table \ref{table:definitions}) are added. Note that all non-lake related climate fields such as vegetation cover or orography were updated in V20 field set comparing to V15 only in relation to the changing lake fields (i.e. if fraction of lake in the grid cell increased then other fractions like vegetation or bare ground should have increased accordingly). \newline

@@

\noindent As focus of this study are lake related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results here was restricted to areas where there have been significant changes in the surface physiographic fields, i.e. a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. Choice of ≥ 10 \% cut-off was adopted as it proofed to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected even though with more severe changes, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise. Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
\end{itemize}
The VESPER models is trained on 2016 data over the entire globe, and then used to predict LST for 2019. The training of a neural network is inherently stochastic - the same model trained twice with the same data can settle in different local optima and so make different predictions. To make our conclusions robust against this training noise, each VESPER model is in turn trained 4 times. For each MODIS ground truth we then have 4 LST predictions per model. We will frequency refer to a ``training noise" as the standard deviation $\sigma$ in the VESPER predictions for the same input fields i.e. each VESPER model will have a corresponding training noise $\sigma_{\rm VM}$. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) we compare the mean absolute error (MAE) between different VESPER models using the a simple metric $\delta_{\rm VM}$ is used:
\begin{equation} 
	\delta_{\rm VM} = \text{MAE}_{\rm VESPER\_VM} - \text{MAE}_{\rm VESPER\_V15} 
\end{equation}
where $M$ represents one of the field set versions V20, V20X or V15X, and $\text{MAE}$ is computed over the whole prediction period of 2019. In turn, the MAE is the error between the prediction of a VESPER model and the Aqua-MODIS LST. A negative $\delta_{\rm VM}$ indicates that the VESPER\_VM LST prediction is more accurate than the  VESPER\_V15 prediction, and vice versa.

@@

\subsection{Evaluation of updated lake fields}\label{sec:3}
To understand if there is a way to automatically and rapidly assess accuracy of updated and/or new surface physiography fields and if their use in the atmospheric model increase predictability a neural network regression model VESPER was trained over 2016 data and used to predict LST for 2019. Generally VESPER’s training noise was confirmed to be  smaller than differences in LST predictions by different VESPER configurations so changes in LST predictability can be meaningfully attributed to the changes in surface physiographic fields. Particular situations where the training noise becomes significant are discussed below. \newline

As a first attempt lake related information is assessed, namely lake cover (and land-sea mask and glacier cover as they are used for lake cover generation) and lake mean depth, that were created from scratch using new up-to-date high-resolution input datasets (see Table \ref{tab:datasources}) for the V20 (and V20X) field set; other surface physiographic fields (see Table \ref{table:definitions}) were regenerated from the same input sources as in initial V15 field set but taking into account that lake related fields were changed. In cases when existing in V15 lake cover water was removed in V20, its place could fill high or low vegetation, glacier or bare ground. To analyse results grid-cells were divided into four categories, and each was investigated in detail (see Table \ref{tab:categorisation} for the results aggregated over the whole globe).

\begin{table*}
	\begin{tabular}{ccccccccc}
		\toprule
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{Training noise of VESPER\_VM, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$, $K$} \\  
		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\
		\hline 
		Lake&1631 & 0.07& 0.02& 0.02& 0.02&  -0.2 & -0.37&-0.37 \\
		Lake-Ground&546 & 0.15 &0.05& 0.04& 0.06 & -0.56 &-0.83& -0.84 \\
		Vegetation&58 & 0.04 &0.1& 0.15& 0.21&  -$4 \times 10^{-3}$&0.04 & -3 $\times 10^{-4}$ \\
		Glacier&1057 & 0.03& 0.08& 0.02& 0.06 & -0.01& -0.22& -0.28  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences VM between mean absolute error (MAE) of [VESPER\_VM LST predictions minus Aqua-MODIS LST] and MAE of [VESPER\_V15 LST predictions minus Aqua-MODIS LST] for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative VM values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive VM values indicate decrease of LST predictability and suggest the presence of erroneous information in the surface physiographic fields. Training noise values are much smaller than the variance between different VESPER’s configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields. The quoted noise is the standard deviation of the prediction errors of Fig. \ref{fig:global_shift_plot}.}
	\label{tab:categorisation}
\end{table*}	
\begin{figure}
	\includegraphics[width=\columnwidth]{global_shift_plot}
	\caption{Distribution of prediction errors in the LST, for each of the 4 grid point categories, for each iteration of V15 and V20. For Lake, Lake-ground and Glacier categories the improvement in V20 relative to V15 is much greater than the intrinsic model noise, with all V20 predictions outperforming all V15 predictions. For the Vegetation category the predictions of V15 and V20 are much more noisy and it is difficult to draw any conclusions for the category as a whole.} 
	\label{fig:global_shift_plot}
\end{figure}

\subsubsection{Category: Lake updates}
\begin{figure}
	\includegraphics[width=\columnwidth]{lakes_histogram_plot}
	\caption{Distribution of prediction errors in the LST for all grid points in the lakes category for V15 and V20.  Each prediction errors is in turn the average of 4 trained iterations of the VESPER model.} 
	\label{fig:example_figure_histogram}
\end{figure}

Lake Updates category shows significant improvements in LST predictability if using V20 field set instead of V15 – prediction accuracy increased globally (over 1631 grid-cells) on average by 0.37K. For the lakes category, the training noise in V20 was generally small $\sigma_{\rm V20} \sim 0.02$ K, with the V15 predictions a little more noisy with $\sigma_{\rm V15} \sim 0.07$ K, but this noise is much less than the improvement - as can be seen in Fig. \ref{fig:global_shift_plot} every V20 iteration significantly outperforms every V15 iteration. In Fig. \ref{fig:example_figure_histogram} we plot the distribution of the mean LST error (averaged across each of the 4 trained VESPER iterations) for all lake grid points, for both V15 and V20. Evidently the V20 field significantly improve the high tail behaviour relative to V15, as well as shifting the median of the distribution to lower errors. Particular regions where the V20 physiographic fields notably improved performance were in Australia and the Aral sea (Fig. \ref{fig:bitstring_100110}). These are two major regions where ephemeral lakes were removed and inland water distribution made up-to-date respectively, as discussed in Surface physiographic fields section. In addition to the areas with a notable improvement in the prediction accuracy, there are some noteworthy regions where the predictions got worse (see red points in Figure \ref{fig:bitstring_100110}) suggesting inaccuracies or lack of information in the updated surface physiographic fields. A few of the most noteworthy grid-cells (see red points highlighted with green circles in Figure \ref{fig:bitstring_100110}) are:
\begin{itemize}

		\item \textbf{Northern India.} Grid-cell lies in the state of Gujarat, India, close to the border with Pakistan. Lake fraction was increased from 0.59 in V15 to 0.71 in V20 field set, along with the lake depth increase from 2.58m to 3.76m. However, this point lies on a river delta within the Great Raan of Kutch, a large area of salt marshes (see Figure \ref{fig:lakes5}), known for having highly seasonal rainfall, with frequent flooding during the monsoon season and a long dry season. The surface itself also undulates with areas of higher sandy ground known as medaks, with greater levels of vegetation. It is evidently a complex and highly time variable area and additional static fraction of fresh water provided via V20 field set is not enough.

	\item \textbf{Salt Lake City, North America.} Grid-cell lies within the Great Salt Lake Desert, just to the west of the Great Salt Lake, Utah, US. Lake fraction was completely removed from over 0.50 in V15 to 0.00 in V20 field set, grid-cell is fully covered with bare ground in V20 field set. Whilst the training noise is significant at this grid point (0.26 for V15, 0.92 for V20) is it smaller than the error metric ($\delta_{V20} = 2.92K$) and we can see from Fig \ref{fig:lake_sub_categories} that the V20 predictions consistently underperform the V15 predictions. Whilst this area primarily is bare ground, satellite imagery also suggests the presence of a presumably highly saline lake (see Figure \ref{fig:lakes3}); in addition area has a large degree of orography and high elevation (∼1300m) which probably further complicates the surface temperature response. A more accurate description that accounts for the seasonality of the surface water and the salinity is necessary here.

	\item \textbf{Tanzania}. There are two grid-cells at the centre and northern edge of Lake Natron, in Tanzania (lies to the south-east of Lake Victoria) where VESPERV20 LST predictions are less accurate than VESPERV15. For the central point (see Figure \ref{fig:lakes1}) the lake fraction was increased from 0.04 in V15 to 0.39 in V20 field set; for the northern edge point lake fraction was also increased in V20 comparing to V15 field set along with a small decrease (∼0.1) in the low vegetation fraction. However, Lake Natron is a highly saline lake that often dries out, with high temperatures, high levels of evaporation and irregular rainfall. It is a highly complex and variable regime that is not well described by simply increasing fraction of permanent fresh water, and indeed results suggest that with current lake parametrization scheme it may be beneficial to keep the lake fraction low or introduce extra descriptor, e.g. salinity.

	\item \textbf{Algeria}. This grid point lies in Algeria, at the northern edge of the Chott Felrhir, an endoheric salt lake. Similar to the Great Salt Lake Desert, lake fraction was completely removed from 0.33 in V15 to 0.0 in V20. However, Chott Felrhir goes through frequent periods of flooding where the lake is filled by multiple large wadi, and corresponding dry periods where the lake becomes a salt pan. As with the Great Salt Lake Desert it is also a highly variable, complex area requires additional consideration of the salinity and the seasonality

	\item \textbf{Lake Chad}

	\item \textbf{Al Fashaga}

	\item \textbf{Tersakan Lake}

	\item \textbf{Lake Urmia}

@@

\textcolor{red}{TK: I have remove discussion of Canada and Caspian. These areas don't show as severe underperformance now - highly time variable}

@@

\noindent Lake-Ground Updates sub-category, which restricts analysis to only points with no significant change in the vegetation, allows to more clearly see the effect of adding/removing water on/from bare ground. Sub-category shows even bigger improvements in LST predictability if using V20 field set instead of V15 (see Table 4) – prediction accuracy increased globally (over 546 grid-cells) on average by 1.12K (training noise in VESPERV15 XX, and in VESPERV20 XX). This indicates that whilst the updated lake fields are globally accurate and informative, providing on average over the globe, over a year, more than an extra Kelvin of predictive performance, the updates to the vegetation fields tamper this performance gain, indicating potential problem with the vegetation fields

>>>>

\subsubsection{Surface physiographic fields}\label{sec:surface_physio}

@@

To expand V15 and V20 lake description (to V15X and V20X respectively, see Table \ref{tab:datasources}) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) would really benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant changes in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover GLDBv3 salt lake list was used. First, in order to identify separate lakes on ~1km resolution lake cover (here max lake distribution based on monthly varying lake cover is used to ease further application), small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data. Note that all non-lake related climate fields such as vegetation cover or orography were updated in V20 field set compared to V15 only in relation to the changing lake fields (i.e. if fraction of lake in the grid cell increased then other fractions like vegetation or bare ground should have increased accordingly).

@@

 \noindent VESPER was trained on selected atmospheric and surface model fields from ERA5 (see Table \ref{table:definitions}), certain static version of the surface physiographic fields (see Table \ref{tab:datasources}), and Aqua-MODIS LST for 2016. Once VESPER was fully trained it was used to predict LST over the whole globe for 2019.  Going forward, as a shorthand we will refer to VESPER trained using $VM$ surface physiographic fields as VESPER\_VM. For instance, a VESPER model trained using the surface physiographic fields from V15, we refer to as VESPER\_V15 (see Table \ref{tab:vesper_table} for an explicit definition of all the VESPER models). All VESPER versions are trained with ERA5 fields for 2016 and with main surface physiographic fields V15. Then depending on the version some or all additional surface physiographic fields (see Table \ref{table:definitions}) are added. VESPER’s predictions can be compared to the initial ERA5 skin temperatures and actual Aqua-MODIS LST for 2019 (see Figure \ref{fig:example_model}. Figure \ref{fig:example_model} shows the mean absolute errors (MAE) globally in the VESPER\_V15 LST predictions, relative to the Aqua-MODIS LST along with the corresponding MAE in the predicted skin temperature from ERA5. We can see that VESPER\_V15 was able to learn corrections to ERA5, especially in the Himalayas and sub-Saharan Africa as well as Australia and the Amazon basin, leading to the globally averaged MAE reduction for predicted LST; the MAE relative to  Aqua-MODIS LST, averaged over all grid points, was 3.9K for ERA5 and 3.0K for VESPER\_V15. \newline

@@

\noindent As the focus of this study is lake-related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results was restricted to areas where there have been significant changes in the surface lake physiographic fields. By ``significant change" we mean a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); for example if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. The choice of ≥ 10 \% as a significance cut-off was adopted as it proved to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected, albeit with more severe changes to their surface fields, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise (training noise is discussed below). Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
The VESPER model is trained on 2016 data over the entire globe, and then used to predict LST for 2019. The training of a neural network is inherently stochastic - the same model trained twice with the same data can settle in different local optima and so make different predictions. To make our conclusions robust against this training noise, each VESPER model is in turn trained 4 times. For each MODIS ground truth we then have 4 LST predictions per model. We define the training noise as the standard deviation $\sigma$ in the VESPER predictions for the same input fields i.e. each VESPER\_VX model will have a corresponding training noise $\sigma_{\rm VM}$. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) we compare the mean absolute error (MAE) between different VESPER models using the a simple metric $\delta_{\rm VM}$:
\begin{equation} 
	\delta_{\rm VM} = \text{MAE}_{\rm VESPER\_VM} - \text{MAE}_{\rm VESPER\_V15} 
\end{equation}
where $M$ represents one of the field set versions V20, V20X or V15X, and $\text{MAE}$ is computed over the whole prediction period of 2019. In turn, the MAE is the error between the prediction of a VESPER model and the Aqua-MODIS LST. A negative $\delta_{\rm VM}$ then indicates that the VESPER\_VM LST prediction is more accurate than the  VESPER\_V15 prediction, and vice versa.

@@

\subsection{Evaluation of updated lake fields}\label{sec:3}
To understand if there is a way to automatically and rapidly assess the accuracy of updated and/or new surface physiography fields, and if their use in the atmospheric model increase predictability, we can compare the prediction accuracy of different VESPER\_VM models. Generally VESPER’s training noise was confirmed to be smaller than differences in LST predictions by different VESPER configurations, so changes in LST predictability can be meaningfully attributed to the changes in surface physiographic fields. Particular situations where the training noise becomes significant are discussed below. \newline

As a first attempt lake-related information is assessed, namely lake cover (and land-sea mask and glacier cover as they are used for lake cover generation) and lake mean depth, that were created from scratch using new up-to-date high-resolution input datasets (see Table \ref{tab:datasources}) for the V20 (and V20X) field set; other surface physiographic fields (see Table \ref{table:definitions}) were regenerated from the same input sources as in initial V15 field set but taking into account that lake related fields were changed. In cases when existing in V15 lake cover water was removed in V20, its place could fill high or low vegetation, glacier or bare ground. To analyse results grid-cells were divided into four categories, and each was investigated in detail (see Table \ref{tab:categorisation} for the results aggregated over the whole globe).
\begin{table*}
	\begin{tabular}{ccccccccc}
		\toprule
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{$\sigma_{\rm VM}$, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$, $K$} \\  
		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\
		\hline 
		Lake&1631 & 0.07& 0.02& 0.02& 0.02&  -0.2 & -0.37&-0.37 \\
		Lake-Ground&546 & 0.15 &0.05& 0.04& 0.06 & -0.56 &-0.83& -0.84 \\
		Vegetation&58 & 0.04 &0.1& 0.15& 0.21&  -$4 \times 10^{-3}$&0.04 & -3 $\times 10^{-4}$ \\
		Glacier&1057 & 0.03& 0.08& 0.02& 0.06 & -0.01& -0.22& -0.28  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences $\delta_{\rm VM}$ between mean absolute error (MAE) of VESPER\_VM  and VESPER\_V15 LST for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative $\delta_{\rm VM}$ values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive $\delta_{\rm VM}$ values indicate a decrease in the LST predictability and suggests the presence of erroneous information in the surface physiographic fields. Training noise values are generally much smaller than the variance between different VESPER configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields. The quoted noise is the standard deviation of the prediction errors of Fig. \ref{fig:global_shift_plot}.}
	\label{tab:categorisation}
\end{table*}	
\begin{figure}
	\includegraphics[width=\columnwidth]{global_shift_plot}
	\caption{Distribution of prediction errors in the LST, for each of the 4 grid point categories, for each iteration of V15 and V20. For Lake, Lake-ground and Glacier categories the improvement in V20 relative to V15 is much greater than the intrinsic model noise, with all V20 predictions outperforming all V15 predictions. For the Vegetation category the predictions of V15 and V20 are much more noisy and it is difficult to draw any conclusions for the category as a whole.} 
	\label{fig:global_shift_plot}
\end{figure}

\subsubsection{Category: Lake updates}
\begin{figure}
	\includegraphics[width=\columnwidth]{lakes_histogram_plot}
	\caption{Distribution of prediction errors in the LST for all grid points in the Lake Updates category for VESPER\_V15 and VESPER\_V20.  Each prediction errors is in turn the average of 4 trained iterations of the VESPER model. The predictions of VESPER\_V20 are evidently and improvement over VESPER\_V15.} 
	\label{fig:example_figure_histogram}
\end{figure}
The Lake Updates category shows significant improvements in LST predictability if using V20 field set instead of V15 – prediction accuracy increased globally (over 1631 grid-cells) on average by 0.37K. For the lakes category, the training noise in V20 was generally small $\sigma_{\rm V20} \sim 0.02$ K, with the V15 predictions a little more noisy with $\sigma_{\rm V15} \sim 0.07$ K, but this noise is much less than the improvement - as can be seen in Fig. \ref{fig:global_shift_plot} every V20 iteration significantly outperforms every V15 iteration. In Fig. \ref{fig:example_figure_histogram} we plot the distribution of the mean LST error (averaged across each of the 4 trained VESPER iterations) for all lake grid points, for both V15 and V20. Evidently the V20 field significantly improve the high tail behaviour relative to V15, as well as shifting the median of the distribution to lower errors. Particular regions where the V20 physiographic fields notably improved performance were in Australia and the Aral sea (e.g. Fig. \ref{fig:bitstring_100110}). These are two major regions where ephemeral lakes were removed and inland water distribution made up-to-date, as discussed in Section \ref{sec:surface_physio}. In addition to the areas with a notable improvement in the prediction accuracy, there are some noteworthy regions where the predictions got worse (see red points in Figure \ref{fig:bitstring_100110}) suggesting inaccuracies or lack of information in the updated surface physiographic fields. A few of the most noteworthy grid-cells (see red points highlighted with green circles in Figure \ref{fig:bitstring_100110} and also Figure \ref{fig:lake_sub_categories}) are:
\begin{itemize}

		\item \textbf{Northern India.} This grid-cell lies in the state of Gujarat, India, close to the border with Pakistan. Here $\delta_{\rm V20} = +4.21$, with $\sigma_{\rm V15} = 2.54$ and $\sigma_{\rm V20}=0.416$. The lake fraction was increased from 0.59 in V15 to 0.71 in V20 field set, along with the lake depth increase from 2.58m to 3.76m.However, this point lies on a river delta within the Great Raan of Kutch, a large area of salt marshes (see Figure \ref{fig:lakes5}), known for having highly seasonal rainfall, with frequent flooding during the monsoon season and a long dry season. The surface itself also undulates with areas of higher sandy ground known as medaks, with greater levels of vegetation. It is evidently a complex and highly time variable area and additional static fraction of fresh water provided via V20 field set is not enough.

	\item \textbf{Salt Lake City, North America.} This grid-cell lies within the Great Salt Lake Desert, just to the west of the Great Salt Lake, Utah, US. Predictions of VESPER\_V20 are worse than VESPER\_V15, with $\delta_{\rm V20} = +2.91$ ( $\sigma_{\rm V15} = 0.26$ and $\sigma_{\rm V20}=0.92$). Whilst the training noise is significant here, it is less than the $\delta_{\rm V20}$ value, and we can see from Fig \ref{fig:lake_sub_categories} that the VESPER\_V20 predictions consistently underperform the VESPER\_V15 predictions. The lake fraction was completely removed from over 0.50 in V15 to 0.00 in V20 field set, meaning that the grid-cell is fully covered with bare ground in V20 field set. Whilst this area primarily is bare ground, satellite imagery also suggests the presence of a presumably highly saline lake (see Figure \ref{fig:lakes3}); in addition area has a large degree of orography and high elevation (∼1300m) which probably further complicates the surface temperature response. A more accurate description that accounts for the seasonality of the surface water and the salinity is necessary here.

	\item \textbf{Tanzania}. There are two grid-cells of interest at the centre and northern edge of Lake Natron, which itself lies to south-east of Lake Victoria, in Tanzania. For both these points VESPER\_V20 predictions are less accurate than VESPER\_V15; for the central point ($\delta_{\rm V20} = +2.45$,$\sigma_{\rm V15} = 0.12$ and $\sigma_{\rm V20}=0.81$, see also Figure \ref{fig:lakes1}) the lake fraction was increased from 0.04 in V15 to 0.39 in V20 field set; for the northern edge point ($\delta_{\rm V20} = +1.57$,$\sigma_{\rm V15} = 0.13$ and $\sigma_{\rm V20}=0.51$) the lake fraction was also increased in V20 comparing to V15 field set along with a small decrease ($\sim$0.1) in the low vegetation fraction. However, Lake Natron is a highly saline lake that often dries out, with high temperatures, high levels of evaporation and irregular rainfall. It is a highly complex and variable regime that is not well described by simply increasing fraction of permanent fresh water, and indeed results suggest that with current lake parametrization scheme it may be beneficial to keep the lake fraction low or introduce extra descriptor, e.g. salinity.

	\item \textbf{Algeria}. This grid point lies in Algeria, at the northern edge of the Chott Felrhir, an endorheic salt lake ($\delta_{\rm V20} = +2.20$,$\sigma_{\rm V15} = 0.41$ and $\sigma_{\rm V20}=0.49$). Similar to the Great Salt Lake Desert, the lake fraction was completely removed from 0.33 in V15 to 0.0 in V20. However, Chott Felrhir goes through frequent periods of flooding where the lake is filled by multiple large wadi, and corresponding dry periods where the lake becomes a salt pan. As with the Great Salt Lake Desert it is also a highly variable, complex area requires additional consideration of the salinity and the seasonality.

	\item \textbf{Lake Chad} This grid point contains Lake Chad, a freshwater endorheic lake in the central part of the Sahel ($\delta_{\rm V20} = +1.74$,$\sigma_{\rm V15} = 0.33$ and $\sigma_{\rm V20}=0.98$). Here the lake fraction was modestly reduced from 0.63 to 0.47. However, Lake Chad is again a highly time variable regime with seasonal droughts and wet seasons. It is a marshy wetland area but the vegetation fractions in both V15 and V20 here are zero. Satellite imagery also shows a large fraction of the surface covered by water and vegetation (Figure \ref{fig:lakes11}).

	\item \textbf{Al Fashaga} This grid point lies in a disputed region between Sudan and Ethiopia called Al Fashaga, close to a tributary of the Nile ($\delta_{\rm V20} = +0.94$,$\sigma_{\rm V15} = 0.14$ and $\sigma_{\rm V20}=0.29$). The updated V20 fields increased the lake fraction at this point from $0$ to $0.14$. The grid cell contains the Upper Atbara and Setit Dam Complex. However, the dam was only recently completed in 2018 - during the training period the damn was still under construction. Consequently whilst the V20 field may be more accurate at the current time, during the period the model was training the V15 field was more accurate, since the damn was not yet built.

	\item \textbf{Lake Tuz}. This grid cell contains a large fraction of Lake Tuz as well as the smaller Lake Tersakan, saline lakes in central Turkey ($\delta_{\rm V20} = +0.85$,$\sigma_{\rm V15} = 0.25$ and $\sigma_{\rm V20}=0.34$). Here the updated physiographic field effectively removed all lake water, with the lake fraction decreasing from 0.14 to 0.005. Whilst the lake is shallow and does dry out in the summer, there is also a large fraction of surface water present (e.g. Fig \ref{fig:lakes44s}) and it is an over correction to completely remove all lake water at this point.

	\item \textbf{Lake Urmia}. This grid cell contains Lake Urmia, which is another saline lake in Iran ($\delta_{\rm V20} = +0.81$,$\sigma_{\rm V15} = 0.12$ and $\sigma_{\rm V20}=0.73$). The updated physiographic fields decreased the lake fraction at this point from 0.77 to 0.39. This was in response to the shrinking of Lake Urmia due to long-timescale droughts and the damming of rivers in Iran. However, this drought broke in 2019 and Lake Urmia is now increasing in size again - satellite imagery now shows a large fraction of the grid cell covered by water (Figure \ref{fig:lakes444s}).
\end{itemize}

@@

@@

\noindent The Lake-Ground Updates sub-category, which restricts analysis to only points with no significant change in the vegetation, allows to more clearly see the effect of adding/removing water on/from bare ground. This sub-category shows even larger improvements in LST predictability if using V20 field set instead of V15 (see Table 4) – prediction accuracy increased globally (over 546 grid-cells) on average by 0.83K ($\sigma_{\rm V15} = 0.15$ and $\sigma_{\rm V20}=0.04$, see also Figure \ref{fig:global_shift_plot}). This indicates that whilst the updated lake fields are globally accurate and informative, providing on average over the globe, over a year, nearly an extra Kelvin of predictive performance, the updates to the vegetation fields tamper this performance gain, indicating potential problem with the vegetation fields

%#ake_haver.png
\begin{figure*}[t]
	\includegraphics[width=0.98\textwidth]{updated_lakes_plot}
	\caption{Differences in the prediction error MAE, between VESPER\_V20 and VESPER\_V15, $\delta_{\rm V20}$,for 2019 at ~31km resolution for ‘Lake Updates’ category (i.e. where lake cover changed significantly). Generally, VESPER\_V20 LST predictions are more accurate, for example in the Aral sea and Australia, indicating that V20 field set is informative and accurate. Particular points where VESPER\_V20 LST prediction gets notably worse compared to VESPER\_V15 are highlighted with green circles and discussed in the text.}
	\label{fig:bitstring_100110}
\end{figure*}
